---
title: "EHR Data QC Pipeline"
output: html_document
date: "`r Sys.Date()`"
---

This script generates tables and figures for EHR data quality control (QC). It processes NLP and codified datasets, ensuring data consistency and reliability for analysis. 

### Requirements

For each sample or institution, you should have two datasets (NLP & codified) that include at least:

- `patient_num`: Character variable for patient ID
- `feature_id`: Character variable for feature (code) ID
- `start_date`: Date of each `feature_id` (in YYYY-MM-DD or YYYY format)

Additionally, you need a data dictionary containing:

- `feature_id`: Character variable for feature (code) ID
- `description`: Text description of `feature_id`

### Setup

Define the following in **Module 1**:

- Directory paths to dictionaries (general and ONCE), NLP, and codified data
- `target_code`: PheCode of primary interest in your study
- `target_cui`: Corresponding CUI of primary interest in your study

*Note that if you are not running this code on O2, you will need to download codified and NLP features from the [ONCE webapp](https://shiny.parse-health.org/ONCE/) and manually specify directory paths for these dictionaries.*

```{r libraries, include=FALSE}

# rm(list = ls())

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, fig.width = 20, fig.height = 8)

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(ggtext)
library(knitr)
library(lubridate)
library(cowplot)
library(RColorBrewer)
library(purrr)

source("Helper_Functions_04152025.R")

```

## Module 1: Data and Dictionary Import & Preparation

This module imports and prepares the NLP and codified datasets for analysis. It also imports three data dictionaries - an institution-specific data dictionary with codified feature descriptions (user-defined) and two ONCE dictionaries for selecting similar features to the target PheCode and CUI (automatically uploaded from O2).

### Outputs

1. Cleaned NLP and codified datasets
2. Filtered data dictionary with target and common code feature descriptions
3. ONCE dictionaries with selected codified and NLP features

```{r module1a}

# ----------------------- USER INPUT -----------------------

# Provide target PheCode and CUI
target_code <- "PheCode:335"
target_cui <- "C0026769"

# If running this script on O2, set to TRUE (otherwise FALSE)
O2 <- FALSE

# Dictionary path
dictionary_path <- "../extdata/fake_dict.csv"

# Data paths
paths <- list(
  nlp1 = "../extdata/fake_nlp.csv",
  codified1 = "../extdata/fake_codified.csv"
)

# Optional second sample/institution (leave commented out if only plotting a single sample/institution)
paths$nlp2 <- "../extdata/fake_nlp_v2.csv"
paths$codified2 <- "../extdata/fake_codified_v2.csv"

# Customize sample names
sample_labels <- list("1" = "UPMC Sample 1", "2" = "UPMC Sample 2")  

# Manual ONCE paths (if O2 = FALSE)
manual_ONCE_path_code <- "../extdata/ONCE_multiple_sclerosis_codified.csv"
manual_ONCE_path_nlp <- "../extdata/ONCE_multiple_sclerosis_nlp.csv"

```

```{r module1b}

# ----------------------- DO NOT EDIT -----------------------

# Load and clean raw data
data_inputs <- load_input_data(paths)

  # Optional: Preview cleaned data files 

    # Sample 1
    # head(data_inputs[["nlp1"]])
    # head(data_inputs[["codified1"]])

    # Sample 2
    # head(nlp2 <- data_inputs[["nlp2"]])
    # head(codified2 <- data_inputs[["codified2"]]). 

# Load ONCE dictionary
ONCE <- clean_ONCE_data(target_code, O2, manual_ONCE_path_code, manual_ONCE_path_nlp)

# Load filtered dictionary
filtered_dictionary <- filter_dictionary(dictionary_path, target_code, target_cui)

```

## Module 2: Patient & Code Summaries

This module summarizes the NLP and codified datasets, including patient counts and prevalence of the target PheCode and CUI. Patient counts are summarized annually (line plots) and overall (bar plots).

### Outputs

1. Patient counts over time
2. Table of follow-up duration statistics

```{r module2}

# ----------------------- DO NOT EDIT -----------------------

# 2.a. Display total sample sizes
sample_sizes <- lapply(names(data_inputs), function(name) {
  data <- data_inputs[[name]]
  data_type <- ifelse(grepl("nlp", name), "NLP", "Codified")
  sample_key <- ifelse(grepl("2", name), "2", "1")
  sample_label <- sample_labels[[sample_key]]
  data.frame(Sample = sample_label, Dataset = data_type, `Number of Patients` = length(unique(data$patient_num)))
}) %>% bind_rows()

knitr::kable(sample_sizes, caption = "Total Sample Sizes")

# 2.b. NLP: generate summaries and plots
nlp_data_inputs <- data_inputs[grepl("nlp", names(data_inputs))]
nlp_summaries <- generate_all_patient_data(nlp_data_inputs, target_cui, sample_labels)

plot_nlp <- generate_line_plot(nlp_summaries$counts, "Sample Size by Year (NLP)")
plot_bar_nlp <- generate_bar_plot(nlp_summaries$summary, "Total Patients vs. Target CUI")

# 2.c. Codified: generate summaries and plots
cod_data_inputs <- data_inputs[grepl("codified", names(data_inputs))]
cod_summaries <- generate_all_patient_data(cod_data_inputs, target_code, sample_labels)

plot_codified <- generate_line_plot(cod_summaries$counts, "Sample Size by Year (Codified)")
plot_bar_codified <- generate_bar_plot(cod_summaries$summary, "Total Patients vs. Target Code")

# 2.d. Extract legends
legend_nlp <- cowplot::get_legend(
  plot_nlp + theme(legend.position = "right")
)

legend_codified <- cowplot::get_legend(
  plot_codified + theme(legend.position = "right")
)

# 2.e. Arrange the plots with legend to the right
combined_plot_nlp <- cowplot::plot_grid(
  plot_nlp + theme(legend.position = "none"),
  plot_bar_nlp,
  legend_nlp,
  ncol = 3,
  rel_widths = c(3, 2, 1)
)

combined_plot_codified <- cowplot::plot_grid(
  plot_codified + theme(legend.position = "none"),
  plot_bar_codified,
  legend_codified,
  ncol = 3,
  rel_widths = c(3, 2, 1)
)

# 2.f. Display plots
print(combined_plot_nlp)
print(combined_plot_codified)

```

## Module 3: Common PheCode Trends (Parent-Child Correlation)

This module examines hierarchical relationships between PheCodes, tracking trends over time for parent and child PheCodes.

### Outputs

1. Plot of rates over time for 3 common PheCodes
2. Plot of patient counts over time for 3 common PheCodes

```{r module3}

# ----------------------- DO NOT EDIT ----------------------- 

# 3.a. Prepare codified data list using sample_labels
codified_data_list <- list()
if (!is.null(data_inputs$codified1)) codified_data_list[["1"]] <- data_inputs$codified1
if (!is.null(data_inputs$codified2)) codified_data_list[["2"]] <- data_inputs$codified2

# 3.b. Generate PheCode summaries
phecode_data <- lapply(codified_data_list, generate_phecode_summary)

phecodes <- c("411", "250", "296")

# 3.c. Loop through PheCodes and generate plots
plots <- lapply(phecodes, function(p) {
  rate_plot <- plot_trends_v2(phecode_data, p, filtered_dictionary, "Rate", "Rate", "Rates", sample_labels)
  count_plot <- plot_trends_v2(phecode_data, p, filtered_dictionary, "Patients", "Patients per Year", "Patient Counts", sample_labels)
  bar_plot <- plot_total_patients_v2(codified_data_list, p, filtered_dictionary, sample_labels)

  combined_plot <- cowplot::plot_grid(
                    count_plot + theme(legend.position = "none"),
                    bar_plot + theme(legend.position = "none"),
                    ncol = 2,
                    rel_widths = c(3, 2))

  list(rate_plot = rate_plot, combined_plot = combined_plot)
})

# 3.d. Print plots
for (i in seq_along(plots)) {
  print(plots[[i]]$rate_plot)
  print(plots[[i]]$combined_plot)
}

```

*Notes: (1) Rates are calculated as the number of patients with the target PheCode per calendar year divided by the total number of patients with any code in the same year.*

## Module 4: Target PheCode-CUI Trends

This module analyzes the relationship between a target PheCode and CUI, including trends in annual rates, patient counts, and intra-patient correlation.

### Outputs

1. Plot of rates over time for target PheCode-CUI pair
2. Plot of patient counts over time for target PheCode-CUI pair
3. Plot of intra-patient correlation over time for target PheCode-CUI pair

```{r module4}

# ----------------------- DO NOT EDIT ----------------------- 

# 4.a. Build codified and NLP lists from sample labels
codified_list <- list()
nlp_list <- list()
if (!is.null(data_inputs$codified1)) codified_list[["1"]] <- data_inputs$codified1
if (!is.null(data_inputs$codified2)) codified_list[["2"]] <- data_inputs$codified2
if (!is.null(data_inputs$nlp1)) nlp_list[["1"]] <- data_inputs$nlp1
if (!is.null(data_inputs$nlp2)) nlp_list[["2"]] <- data_inputs$nlp2

# 4.b. Summarize across codified and NLP
summarized <- summarize_target_code_trends(
  codified_list = codified_list,
  nlp_list = nlp_list,
  target_code = target_code,
  target_cui = target_cui,
  sample_labels = sample_labels 
)

# 4.c. Plot trends across samples/institutions
trend_plots <- plot_target_code_trends(
  summary_data = summarized$combined,
  target_code = target_code,
  target_cui = target_cui,
  dictionary_code = ONCE$code,
  dictionary_nlp = ONCE$nlp
)

# 4.d. Plot correlation across samples/institutions
correlation_plot <- plot_target_code_correlation(
  wide_data = summarized$wide,
  target_code = target_code,
  target_cui = target_cui,
  summary_data = summarized$combined,
  dictionary_code = ONCE$code,
  dictionary_nlp = ONCE$nlp
)

# 4.e. Print plots
trend_plots$rates
trend_plots$counts
correlation_plot

```

*Notes: (1) Rates are calculated as the number of patients with the target code per calendar year divided by the total number of patients with any code in the same year; (2) Intra-patient correlations are calculated as the Spearman correlation between the code counts for patients in the same year*

## Module 5: ONCE Feature Trends (Diagnoses, Medications, Labs, Procedures)

This module identifies the top five related codes in different categories (diagnosis, medication, lab, procedure, CUI) based on ONCE feature similarity and tracks their trends over time. Patient counts are summarized annually (line plots) and overall (bar plots).

### Outputs

1. Plot of rates over time for related features
2. Plot of patient counts over time for related features

```{r module5}

# ----------------------- DO NOT EDIT -----------------------

# 5.a. Build codified and NLP lists from data_inputs
codified_list <- list()
nlp_list <- list()
if (!is.null(data_inputs$codified1)) codified_list[["1"]] <- data_inputs$codified1
if (!is.null(data_inputs$codified2)) codified_list[["2"]] <- data_inputs$codified2
if (!is.null(data_inputs$nlp1)) nlp_list[["1"]] <- data_inputs$nlp1
if (!is.null(data_inputs$nlp2)) nlp_list[["2"]] <- data_inputs$nlp2

types <- c("Diagnosis", "Medication", "Lab", "Procedure", "CUI")

# 5.b. Generate and display plots
plot_list <- lapply(types, function(t) {
  result <- plot_related_code_trends_v2(
    Type = t,
    target_code = target_code,
    target_cui = target_cui,
    codified_list = codified_list,
    nlp_list = nlp_list,
    sample_labels = sample_labels,
    ONCE = ONCE
  )

  if (!is.null(result[[1]])) {
    print(result[[1]][[1]])  # Line plot (rates)
    print(result[[1]][[2]])  # Line + bar combined plot
  } else {
    message(result[[2]])
  }
})

```


