---
title: "EHR Data QC Pipeline"
output: html_document
date: "`r Sys.Date()`"
---

This script generates tables and figures for EHR data quality control (QC). It processes NLP and codified datasets, ensuring data consistency and reliability for analysis.

### Requirements

You should have two datasets (NLP & codified) that include at least:

- `patient_num`: Character variable for patient ID
- `feature_id`: Character variable for feature (code) ID
- `start_date`: Date of each `feature_id`

Additionally, you need a data dictionary containing:

- `feature_id`: Character variable for feature (code) ID
- `description`: Text description of `feature_id`

### Setup

Define the following in **Module 1**:

- Directory paths to dictionary, NLP, and codified data
- `target_code`: PheCode of primary interest in your study
- `target_cui`: Corresponding CUI of primary interest in your study

Note that if you are not running this code on O2, you will need to download codified and NLP features from the [ONCE webapp](https://shiny.parse-health.org/ONCE/) and manually specify directory paths for these dictionaries.

```{r libraries, include=FALSE}

# rm(list = ls())

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, fig.width = 20, fig.height = 8)

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(ggtext)
library(knitr)
library(lubridate)
library(cowplot)
library(RColorBrewer)

source("Helper_Functions_02262025.R")

theme_global <- theme_minimal() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 18),
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_textbox_simple(size = 16, lineheight = 1.3),  
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

theme_bar <- theme_minimal() +
  theme(
    axis.text = element_text(size = 18),
    axis.title = element_text(size = 18),
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_textbox_simple(size = 16, lineheight = 1.3),  
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

set_palette <- scale_color_brewer(palette = "Set2")
set_fill_palette <- scale_fill_brewer(palette = "Set2")

```

## Module 0: Sample QC Data

This optional module samples 1,000 unique patients from the intersection of the NLP and codified datasets to speed up QC processing.

### Outputs

1. Sampled NLP dataset (to be used in Module 1)
2. Sampled codified dataset (to be used in Module 1)

```{r module0, include=FALSE}

# nlp_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/MS/data_table/data_processed/UPMC_MS_2011_to_2021_NLP_processed_data_2023-10-21.csv"
# codified_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/MS/data_table/data_processed/UPMC_MS_2004_to_2022_Codified_processed_data_2023-08-25.csv"

# nlp <- read.csv(nlp_path)
# codified <- read.csv(codified_path)

# set.seed(123) 
# selected_patients <- sample(intersect(unique(nlp$patient_num), unique(codified$patient_num)), 1000)
 
# nlp_sample <- nlp[nlp$patient_num %in% selected_patients, ]
# codified_sample <- codified[codified$patient_num %in% selected_patients, ]

## Save sampled nlp and codified data to upload in next step
# write.csv(nlp_sample,file="/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/nlp_sample_1000.csv",row.names = FALSE)
# write.csv(codified_sample,file="/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/codified_sample_1000.csv",row.names = FALSE)

```

## Module 1: Data and Dictionary Import & Preparation

This module imports and prepares the NLP and codified datasets for analysis. It also imports three data dictionaries - an institution-specific data dictionary with codified feature descriptions (user-defined) and two ONCE dictionaries for selecting similar features to the target PheCode and CUI (automatically uploaded from O2).

### Outputs

1. Cleaned NLP and codified datasets
2. Filtered data dictionary with target and common code feature descriptions
3. ONCE dictionaries with selected codified and NLP features

```{r module1}

# ----------------------- EDIT ----------------------- 

# Provide directory paths to dictionary, nlp and codified data (csv)

# O2 example (set O2 to TRUE below)
# nlp_path <- "/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/nlp_sample_1000.csv"
# codified_path <- "/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/codified_sample_1000.csv"
# dictionary_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/data_mapping/mapping_clean/UPMC_Codified_data_codebook_with_descriptions_2023-05-15.csv"

# Local example (set O2 to FALSE below)
nlp_path <- "/Users/kimberlygreco/Dropbox/CELEHS/testdata/fake_nlp.csv"
codified_path <- "/Users/kimberlygreco/Dropbox/CELEHS/testdata/fake_codified.csv"
dictionary_path <- "/Users/kimberlygreco/Dropbox/CELEHS/testdata/fake_dict.csv"

# Provide target PheCode and CUI 
target_code <- "PheCode:335"
target_cui <- "C0026769"

# If running in O2, set O2 to TRUE; if not, set O2 to FALSE and provide manual directory paths to ONCE dictionaries for codified and nlp data
O2 <- FALSE
manual_ONCE_path_code <- "/Users/kimberlygreco/Dropbox/CELEHS/testdata/ONCE_multiple_sclerosis_codified.csv"
manual_ONCE_path_nlp <- "/Users/kimberlygreco/Dropbox/CELEHS/testdata/ONCE_multiple_sclerosis_nlp.csv"


# ----------------------- DO NOT EDIT ----------------------- 

# 1.a. Import nlp & codified data
nlp <- read.csv(nlp_path)
codified <- read.csv(codified_path)

codified <- codified %>%
  mutate(
    patient_num = as.character(patient_num),  
    year = format(as.Date(start_date), "%Y"),
    feature_id = str_replace(feature_id, "CCS-PCS", "CCS")
  )

nlp <- nlp %>%
  mutate(
    patient_num = as.character(patient_num), 
    year = format(as.Date(start_date), "%Y"),
    feature_id = str_replace(feature_id, "CCS-PCS", "CCS")
  )

# 1.b. Data dictionary processing
dictionary <- read.csv(dictionary_path)

filtered_dictionary <- dictionary %>%
  filter(
    grepl(target_code, feature_id) |        
    grepl(target_cui, feature_id) |         
    grepl("^PheCode:250(\\.\\d+)?$", feature_id) | 
    grepl("^PheCode:411(\\.\\d+)?$", feature_id) |  
    grepl("^PheCode:296(\\.\\d+)?$", feature_id)  
  ) %>%
  dplyr::select(feature_id, description) %>%
  distinct(feature_id, .keep_all = TRUE) %>%
  mutate(description = case_when(
    str_detect(description, "forms of") ~ str_replace(description, "(.*)forms of.*", "\\1forms"), 
    TRUE ~ str_replace_all(description, "\\s*\\(.*?\\)|,.*", "")
  )) %>%
  mutate(description = tolower(description)) 

# 1.c. ONCE dictionary processing
ONCE <- clean_ONCE_data(target_code, O2, manual_ONCE_path_code, manual_ONCE_path_nlp)

```

## Module 2: Patient, Code, and Follow-up Summaries

This module summarizes the NLP and codified datasets, including patient counts, prevalence of the target PheCode and CUI, and duration of patient follow-up. Patient counts are summarized annually (line plots) and overall (bar plots).

### Outputs

1. Patient counts over time
2. Table of follow-up duration statistics

```{r, module2}

# ----------------------- DO NOT EDIT ----------------------- 

# Compute total unique patients
total_sample_size <- data.frame(
  Dataset = c("NLP", "Codified"),
  "Number of Patients" = c(length(unique(nlp$patient_num)), length(unique(codified$patient_num))),
  check.names = FALSE 
)

knitr::kable(total_sample_size, caption = "Total Sample Size")

# Generate patient counts by year
nlp_patient_counts <- generate_patient_counts(nlp, target_cui)
codified_patient_counts <- generate_patient_counts(codified, target_code)

# Compute total unique patients per dataset
total_patients_nlp <- data.frame(
  Feature = c("Total NLP Patients", "Patients with Target CUI"),
  Total_Patients = c(length(unique(nlp$patient_num)), length(unique(nlp$patient_num[nlp$feature_id == target_cui])))
)

total_patients_codified <- data.frame(
  Feature = c("Total Codified Patients", "Patients with Target PheCode"),
  Total_Patients = c(length(unique(codified$patient_num)), length(unique(codified$patient_num[codified$feature_id == target_code])))
)

# Generate plots
plot_nlp <- generate_line_plot(nlp_patient_counts, "Sample Size by Year (NLP)")
plot_codified <- generate_line_plot(codified_patient_counts, "Sample Size by Year (Codified)")
plot_bar_nlp <- generate_bar_plot(total_patients_nlp, "Total Patients vs. Target CUI")
plot_bar_codified <- generate_bar_plot(total_patients_codified, "Total Patients vs. Target Code")

# Extract legends
legend_nlp <- cowplot::get_legend(plot_nlp)
legend_codified <- cowplot::get_legend(plot_codified)

# Arrange the plots
combined_plot_nlp <- cowplot::plot_grid(
  plot_nlp + theme(legend.position = "none"), 
  plot_bar_nlp, 
  legend_nlp, 
  ncol = 3, rel_widths = c(3, 2, 1)
)

combined_plot_codified <- cowplot::plot_grid(
  plot_codified + theme(legend.position = "none"), 
  plot_bar_codified, 
  legend_codified, 
  ncol = 3, rel_widths = c(3, 2, 1)
)

# Display both plots
print(combined_plot_nlp)
print(combined_plot_codified)

```

```{r followup, include=FALSE, eval=FALSE}

# Follow-up period statistics
nlp_follow_up <- nlp %>%
  group_by(patient_num) %>%
  summarise(
    first_date = min(as.Date(start_date, format = "%Y-%m-%d")),
    last_date = max(as.Date(start_date, format = "%Y-%m-%d")),
    follow_up_years = as.numeric(difftime(last_date, first_date, units = "days")) / 365.25) %>%
  mutate(
    first_date = format(first_date, "%b %Y"),
    last_date = format(last_date, "%b %Y"))
 
codified_follow_up <- codified %>%
  group_by(patient_num) %>%
  summarise(
    first_date = min(as.Date(start_date, format = "%Y-%m-%d")),
    last_date = max(as.Date(start_date, format = "%Y-%m-%d")),
    follow_up_years = as.numeric(difftime(last_date, first_date, units = "days")) / 365.25) %>%
  mutate(
    first_date = format(first_date, "%b %Y"),
    last_date = format(last_date, "%b %Y"))

nlp_stats <- nlp_follow_up %>%
  summarise(
    mean_follow_up = round(mean(follow_up_years),2),
    median_follow_up = round(median(follow_up_years),2),
    min_follow_up = round(min(follow_up_years),2),
    max_follow_up = round(max(follow_up_years),2),
    first_overall_date = min(first_date),
    last_overall_date = max(last_date)
  )

codified_stats <- codified_follow_up %>%
  summarise(
    mean_follow_up = round(mean(follow_up_years),2),
    median_follow_up = round(median(follow_up_years),2),
    min_follow_up = round(min(follow_up_years),2),
    max_follow_up = round(max(follow_up_years),2),
    first_overall_date = min(first_date),
    last_overall_date = max(last_date)
  )

nlp_stats_df <- nlp_stats %>%
  rename(`Mean Follow-up (Years)` = mean_follow_up,
         `Median Follow-up (Years)` = median_follow_up,
         `Min Follow-up (Years)` = min_follow_up,
         `Max Follow-up (Years)` = max_follow_up,
         `First Code Date` = first_overall_date,
         `Last Code Date` = last_overall_date)

codified_stats_df <- codified_stats %>%
  rename(`Mean Follow-up (Years)` = mean_follow_up,
         `Median Follow-up (Years)` = median_follow_up,
         `Min Follow-up (Years)` = min_follow_up,
         `Max Follow-up (Years)` = max_follow_up,
         `First Code Date` = first_overall_date,
         `Last Code Date` = last_overall_date)

knitr::kable(nlp_stats_df, caption = "NLP Follow-up Statistics")
knitr::kable(codified_stats_df, caption = "Codified Follow-up Statistics")

```

## Module 3: Common PheCode Trends (Parent-Child Correlation)

This module examines hierarchical relationships between PheCodes, tracking trends over time for parent and child PheCodes.

### Outputs

1. Plot of rates over time for 3 common PheCodes
2. Plot of patient counts over time for 3 common PheCodes

```{r module3}

# ----------------------- DO NOT EDIT ----------------------- 

# Generate PheCode summary
total_phecode_summary <- generate_phecode_summary(codified)

# Generate and display plots for each PheCode pattern
phecodes <- c("411", "250", "296")

plots <- lapply(phecodes, function(p) {
  rate_plot <- plot_trends(total_phecode_summary, p, filtered_dictionary, "Rate", "Rate", "Rates")
  count_plot <- plot_trends(total_phecode_summary, p, filtered_dictionary, "Patients", "Patients per Year", "Patient Counts")
  bar_plot <- plot_total_patients(codified, p, filtered_dictionary)  
  legend <- cowplot::get_legend(count_plot)
  combined_plot <- cowplot::plot_grid(count_plot + theme(legend.position = "none"), bar_plot, legend, ncol = 3, rel_widths = c(2, .75, 2))
  list(rate_plot = rate_plot, combined_plot = combined_plot)
})

print(plots[[1]]$rate_plot)
print(plots[[1]]$combined_plot)
print(plots[[2]]$rate_plot)
print(plots[[2]]$combined_plot)
print(plots[[3]]$rate_plot)
print(plots[[3]]$combined_plot)

```

*Notes: (1) Rates are calculated as the number of patients with the target PheCode per calendar year divided by the total number of patients with any code in the same year.*

## Module 4: Target PheCode-CUI Trends

This module analyzes the relationship between a target PheCode and CUI, including trends in annual rates, patient counts, and intra-patient correlation.

### Outputs

1. Plot of rates over time for target PheCode-CUI pair
2. Plot of patient counts over time for target PheCode-CUI pair
3. Plot of intra-patient correlation over time for target PheCode-CUI pair

```{r module4}

# ----------------------- DO NOT EDIT ----------------------- 

# Generate data summaries
nlp_target <- nlp[nlp$feature_id == target_cui,]
codified_nlp_target <- rbind(codified,nlp_target)

phecode_cui_wide <- codified_nlp_target %>%
  group_by(year, patient_num, feature_id) %>%
  summarise(Code_Count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = feature_id,
    values_from = Code_Count,
    values_fill = 0
  )

target_phecode_summary <- codified %>%
  filter(str_detect(feature_id, target_code)) %>%
  group_by(year) %>%
  summarise(
    PheCode_Patients = n_distinct(patient_num),
    .groups = "drop"
  )

target_cui_summary <- nlp %>%
  filter(str_detect(feature_id, target_cui)) %>%
  group_by(year) %>%
  summarise(
    CUI_Patients = n_distinct(patient_num),
    .groups = "drop"
  )

target_total_patients <- codified_nlp_target %>%
  group_by(year) %>%
  summarise(Total_Patients = n_distinct(patient_num), .groups = "drop")

target_combined <- target_phecode_summary %>%
  left_join(target_cui_summary, by = "year") %>%
  left_join(target_total_patients, by = "year") %>%
  mutate(
    PheCode_Rate = PheCode_Patients / Total_Patients,
    CUI_Rate = CUI_Patients / Total_Patients
  )

# Extract descriptions from dictionary
phecode_description <- get_phecode_description(gsub("PheCode:", "", target_code), ONCE$code)

cui_description <- ONCE$nlp %>%
  filter(feature_id == target_cui) %>%
  pull(description)
cui_description <- ifelse(length(cui_description) > 0, cui_description, "description not found")

# 4.a. Plot rate for target PheCode-CUI pair
plot_rates_target <- ggplot(target_combined, aes(x = as.numeric(year))) +
  geom_line(aes(y = PheCode_Rate, color = "Target PheCode", group = 1)) +
  geom_line(aes(y = CUI_Rate, color = "Target CUI", group = 2)) +
  geom_point(aes(y = PheCode_Rate, color = "Target PheCode")) +
  geom_point(aes(y = CUI_Rate, color = "Target CUI")) +
  labs(
    title = paste("Rates for Target Codes"),
    subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                       "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
    x = "Year",
    y = "Rate",
    color = "Code"
  ) +
  theme_global +
  set_palette +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 

# 4.b. Plot patient counts for target PheCode-CUI pair
plot_counts_target <- ggplot(target_combined, aes(x = as.numeric(year))) +
  geom_line(aes(y = PheCode_Patients, color = "Target PheCode", group = 1)) +
  geom_line(aes(y = CUI_Patients, color = "Target CUI", group = 2)) +
  geom_point(aes(y = PheCode_Patients, color = "Target PheCode")) +
  geom_point(aes(y = CUI_Patients, color = "Target CUI")) +
  geom_line(
    aes(y = Total_Patients, group = 3),
    linetype = "dotted",
    color = "black"
  ) +  
  labs(
    title = paste("Patient Counts for Target Codes"),
    subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                       "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
    x = "Year",
    y = "Patients",
    color = "Code"
  ) +
  theme_global +
  set_palette +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 

# 4.c. Function to plot correlation for target PheCode-CUI pair
corr_target <- function(target_code, target_cui) {
  
  # Check if relevant columns exist
  if (!target_code %in% colnames(phecode_cui_wide) || !target_cui %in% colnames(phecode_cui_wide)) {
    stop(paste("Not enough data for Target Code:", target_code, "or Target CUI:", target_cui))
  }
  
  # Calculate correlations by year
  correlation_by_year <- phecode_cui_wide %>%
    group_by(year) %>%
    summarise(
      Correlation = {
        # Get data for the current year
        data_for_year <- pick(everything())
        
        # Check for zero standard deviation
        if (sd(data_for_year[[target_code]], na.rm = TRUE) == 0 ||
            sd(data_for_year[[target_cui]], na.rm = TRUE) == 0) {
          NA  
        } else {
          cor(data_for_year[[target_code]], data_for_year[[target_cui]], use = "pairwise.complete.obs", method = "spearman")
        }
      },
      .groups = "drop"
    ) %>%
    filter(!is.na(Correlation))  # Remove rows with NA correlations
  
  # Plot correlations
  ggplot(correlation_by_year, aes(x = as.numeric(year), y = Correlation)) +
    geom_line() +
    geom_point() +
    labs(
      title = "Intra-Patient Correlation for Target Codes",
      subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                        "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
      x = "Year",
      y = "Correlation"
    ) +
    theme_global +
    set_palette +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 
}

plot_corr_target <- corr_target(target_code, target_cui)

plot_rates_target
plot_counts_target
plot_corr_target

```


*Notes: (1) Rates are calculated as the number of patients with the target code per calendar year divided by the total number of patients with any code in the same year; (2) Intra-patient correlations are calculated as the Spearman correlation between the code counts for patients in the same year; (3) Black dotted line represent total patient counts per calendar year (denominator of rate).*

## Module 5: ONCE Feature Trends (Diagnoses, Medications, Labs, Procedures)

This module identifies the top five related codes in different categories (diagnosis, medication, lab, procedure, CUI) based on ONCE feature similarity and tracks their trends over time. Patient counts are summarized annually (line plots) and overall (bar plots).

### Outputs

1. Plot of rates over time for related features
2. Plot of patient counts over time for related features

```{r module5, message=FALSE}

# ----------------------- DO NOT EDIT ----------------------- 

types <- c("Diagnosis", "Medication", "Lab", "Procedure", "CUI")

plot_list <- lapply(types, function(t) {
  result <- plot_related_code_trends(Type = t, target_code, target_cui)
  
  if (!is.null(result[[1]])) {
    print(result[[1]][[1]])  
    print(result[[1]][[2]]) 
  } else {
    message(result[[2]])  
  }
})

```




