
---
title: "EHR Data QC Pipeline"
date: "`r Sys.Date()`"
---

This script generates a set of tables and figures for EHR data QC. To run this script:

You should have 2 datasets (NLP & codified) which contain at least the following variables:

1. `patient_num` - character variable for patient ID
2. `feature_id` - character variable for feature (code) ID
3. `start_date` - date of each feature_id

You should have a data dictionary which contains at least the following variables:

1. `feature_id` - character variable for feature (code) ID
2. `description` - text description of feature_id

You should define the following in **Module 1** (the rest of the script should not be modified):

1. Directory paths to dictionary, NLP and codified data
2. `target_code` - PheCode of primary interest in your study
3. `target_cui` - Corresponding CUI of primary interest in your study

```{r libraries, include=FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)

# rm(list = ls())

library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(ggtext)
library(knitr)
library(lubridate)
library(cowplot)

```

## Module 0 (OPTIONAL): Sample QC Data

This optional module samples 1,000 unique patients from the intersection of the NLP and codified datasets for faster QC processing.  

### Outputs

1. Sampled NLP dataset to save for Module 1
2. Sampled codified dataset to save for Module 1

```{r module0, include=FALSE}

# nlp_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/MS/data_table/data_processed/UPMC_MS_2011_to_2021_NLP_processed_data_2023-10-21.csv"
# codified_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/MS/data_table/data_processed/UPMC_MS_2004_to_2022_Codified_processed_data_2023-08-25.csv"

# nlp <- read.csv(nlp_path)
# codified <- read.csv(codified_path)

# set.seed(123) 
# selected_patients <- sample(intersect(unique(nlp$patient_num), unique(codified$patient_num)), 1000)
 
# nlp_sample <- nlp[nlp$patient_num %in% selected_patients, ]
# codified_sample <- codified[codified$patient_num %in% selected_patients, ]

## Save sampled nlp and codified data to upload in next step
# write.csv(nlp_sample,file="/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/nlp_sample_1000.csv",row.names = FALSE)
# write.csv(codified_sample,file="/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/codified_sample_1000.csv",row.names = FALSE)

```

## Module 1: Data and Dictionary Import & Preparation

This module imports and prepares the NLP and codified datasets for analysis. It also imports two data dictionaries - a general data dictionary with feature descriptions and the ONCE dictionary for selecting similar features to the target PheCode.

### Outputs

1. Cleaned NLP and codified datasets
2. Filtered data dictionary with target and common code feature descriptions
3. ONCE dictionary with selected features

```{r module1}

# ----------------------- EDIT ----------------------- 

# Provide directory paths to dictionary, nlp and codified data (csv)
nlp_path <- "/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/nlp_sample_1000.csv"
codified_path <- "/n/data1/hsph/biostat/celehs/lab/kimgreco/AI Paper/QC script/codified_sample_1000.csv"
dictionary_path <- "/n/data1/hsph/biostat/celehs/lab/SHARE/UPMC/data_mapping/mapping_clean/UPMC_Codified_data_codebook_with_descriptions_2023-05-15.csv"

# Provide target PheCode and CUI 
target_code <- "PheCode:335"
target_cui <- "C0026769"

# ----------------------- DO NOT EDIT ----------------------- 

# 1.a. Import nlp & codified data
nlp <- read.csv(nlp_path)
codified <- read.csv(codified_path)

codified <- codified %>%
  mutate(
    patient_num = as.character(patient_num),  
    year = format(as.Date(start_date), "%Y"),
    feature_id = str_replace(feature_id, "CCS-PCS", "CCS")
  )

nlp <- nlp %>%
  mutate(
    patient_num = as.character(patient_num), 
    year = format(as.Date(start_date), "%Y"),
    feature_id = str_replace(feature_id, "CCS-PCS", "CCS")
  )

# 1.b. Data dictionary processing
dictionary <- read.csv(dictionary_path)

filtered_dictionary <- dictionary %>%
  filter(
    grepl(target_code, feature_id) |        
    grepl(target_cui, feature_id) |         
    grepl("^PheCode:250(\\.\\d+)?$", feature_id) | 
    grepl("^PheCode:411(\\.\\d+)?$", feature_id) |  
    grepl("^PheCode:296(\\.\\d+)?$", feature_id)  
  ) %>%
  dplyr::select(feature_id, description) %>%
  distinct(feature_id, .keep_all = TRUE) %>%
  mutate(description = case_when(
    str_detect(description, "forms of") ~ str_replace(description, "(.*)forms of.*", "\\1forms"), 
    TRUE ~ str_replace_all(description, "\\s*\\(.*?\\)|,.*", "")
  )) %>%
  mutate(description = tolower(description)) 

# 1.c. ONCE dictionary processing

# Dynamic O2 directory path to ONCE dictionary based on target_code
formatted_code <- gsub(":", "", target_code)  
formatted_code <- gsub("\\.", "_", formatted_code)  
ONCE_path <- paste0("/n/data1/hsph/biostat/celehs/lab/sm731/ONCE_features/codified/", formatted_code, "_cod_features.csv")

ONCE_dictionary <- read.csv(ONCE_path)

ONCE_dictionary <- ONCE_dictionary %>%
  dplyr::select(Variable, Description, target_similarity) %>%
  rename(feature_id = Variable, description = Description) %>%
  mutate(description = tolower(description))

```

## Module 2: Patient, Code, and Follow-up Summaries

This module provides summary statistics for the NLP and codified datasets, including patient counts, the prevalence of the target PheCode and CUI, and duration of patient follow-up.

### Outputs

1. Patient counts over time
2. Table of follow-up duration statistics

```{r module2}

# ----------------------- DO NOT EDIT ----------------------- 

options(repr.plot.width = 8, repr.plot.height = 6)

# 2.a. Total sample size (number of unique patients)
total_sample_size <- data.frame(
  Dataset = c("NLP", "Codified"),
  "Number of Patients" = c(length(unique(nlp$patient_num)), length(unique(codified$patient_num))),
  check.names = FALSE 
)

knitr::kable(total_sample_size, caption = "Total Sample Size")

# 2.b. Total sample size BY YEAR
sample_size_by_year <- list(
  nlp = nlp %>%
    group_by(year) %>%
    summarise(total_patients = n_distinct(patient_num)),
  codified = codified %>%
    group_by(year) %>%
    summarise(total_patients = n_distinct(patient_num))
)

# 2.c. Sample size with target disease
patients_with_target_phecode <- unique(codified$patient_num[codified$feature_id == target_code])
patients_with_target_cui <- unique(nlp$patient_num[nlp$feature_id == target_cui])

patients_with_target <- data.frame(
  Dataset = c("PheCode", "CUI"),
  `Number of Patients` = c(length(patients_with_target_phecode), length(patients_with_target_cui)),
  check.names = FALSE
)

knitr::kable(patients_with_target, caption = "Patients with Target Disease")

# 2.d. Sample size with target disease BY YEAR
sample_size_with_target_by_year <- list(
  phecode = codified %>%
    filter(feature_id == target_code) %>%
    group_by(year) %>%
    summarise(target_patients = n_distinct(patient_num)),
  cui = nlp %>%
    filter(feature_id == target_cui) %>%
    group_by(year) %>%
    summarise(target_patients = n_distinct(patient_num))
)

# Merge total sample size and target sample size data
sample_size_by_year$nlp <- sample_size_by_year$nlp %>%
  left_join(sample_size_with_target_by_year$cui, by = "year") %>%
  rename(Year = year, `Total` = total_patients, `Target CUI` = target_patients) %>%
  pivot_longer(cols = c(`Total`, `Target CUI`), names_to = "Type", values_to = "Patients")

sample_size_by_year$codified <- sample_size_by_year$codified %>%
  left_join(sample_size_with_target_by_year$phecode, by = "year") %>%
  rename(Year = year, `Total` = total_patients, `Target PheCode` = target_patients) %>%
  pivot_longer(cols = c(`Total`, `Target PheCode`), names_to = "Type", values_to = "Patients")

# Plot for NLP data
plot_nlp <- ggplot(sample_size_by_year$nlp, aes(x = Year, y = Patients, color = Type, group = Type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Sample Size by Year (NLP)", x = "Year", y = "Number of Patients", color = "Sample") +
  theme_minimal() +
  scale_color_manual(values = c("Total" = "darkblue", "Target CUI" = "lightblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot for Codified data
plot_codified <- ggplot(sample_size_by_year$codified, aes(x = Year, y = Patients, color = Type, group = Type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Sample Size by Year (Codified)", x = "Year", y = "Number of Patients", color = "Sample") +
  theme_minimal() +
  scale_color_manual(values = c("Total" = "darkblue", "Target PheCode" = "lightblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Arrange the two plots side by side
plot_nlp
plot_codified
```

```{r followup, include=FALSE, eval=FALSE}

# 3.e. Follow-up period statistics
nlp_follow_up <- nlp %>%
  group_by(patient_num) %>%
  summarise(
    first_date = min(as.Date(start_date, format = "%Y-%m-%d")),
    last_date = max(as.Date(start_date, format = "%Y-%m-%d")),
    follow_up_years = as.numeric(difftime(last_date, first_date, units = "days")) / 365.25) %>%
  mutate(
    first_date = format(first_date, "%b %Y"),
    last_date = format(last_date, "%b %Y"))
 
codified_follow_up <- codified %>%
  group_by(patient_num) %>%
  summarise(
    first_date = min(as.Date(start_date, format = "%Y-%m-%d")),
    last_date = max(as.Date(start_date, format = "%Y-%m-%d")),
    follow_up_years = as.numeric(difftime(last_date, first_date, units = "days")) / 365.25) %>%
  mutate(
    first_date = format(first_date, "%b %Y"),
    last_date = format(last_date, "%b %Y"))

nlp_stats <- nlp_follow_up %>%
  summarise(
    mean_follow_up = round(mean(follow_up_years),2),
    median_follow_up = round(median(follow_up_years),2),
    min_follow_up = round(min(follow_up_years),2),
    max_follow_up = round(max(follow_up_years),2),
    first_overall_date = min(first_date),
    last_overall_date = max(last_date)
  )

codified_stats <- codified_follow_up %>%
  summarise(
    mean_follow_up = round(mean(follow_up_years),2),
    median_follow_up = round(median(follow_up_years),2),
    min_follow_up = round(min(follow_up_years),2),
    max_follow_up = round(max(follow_up_years),2),
    first_overall_date = min(first_date),
    last_overall_date = max(last_date)
  )

nlp_stats_df <- nlp_stats %>%
  rename(`Mean Follow-up (Years)` = mean_follow_up,
         `Median Follow-up (Years)` = median_follow_up,
         `Min Follow-up (Years)` = min_follow_up,
         `Max Follow-up (Years)` = max_follow_up,
         `First Code Date` = first_overall_date,
         `Last Code Date` = last_overall_date)

codified_stats_df <- codified_stats %>%
  rename(`Mean Follow-up (Years)` = mean_follow_up,
         `Median Follow-up (Years)` = median_follow_up,
         `Min Follow-up (Years)` = min_follow_up,
         `Max Follow-up (Years)` = max_follow_up,
         `First Code Date` = first_overall_date,
         `Last Code Date` = last_overall_date)

knitr::kable(nlp_stats_df, caption = "NLP Follow-up Statistics")
knitr::kable(codified_stats_df, caption = "Codified Follow-up Statistics")

```

## Module 3: Common PheCode Trends (Parent-Child Correlation)

This module examines hierarchical relationships between PheCodes, tracking trends over time for parent and child PheCodes. It calculates annual rates and patient counts for each PheCode level.

### Outputs

1. Plot of rates over time for 3 common PheCodes
2. Plot of patient counts over time for 3 common PheCodes

```{r module3}

# ----------------------- DO NOT EDIT -----------------------

options(repr.plot.width = 8, repr.plot.height = 6)

# Generate data summaries
phecode_summary <- codified %>%
  group_by(year, feature_id) %>%
  summarise(Count = n(), Patients = n_distinct(patient_num), .groups = "drop") %>%
  arrange(year, feature_id)

total_patients_per_year <- codified %>%
  group_by(year) %>%
  summarise(Total_Patients = n_distinct(patient_num), .groups = "drop")

phecode_summary <- phecode_summary %>%
  left_join(total_patients_per_year, by = "year") %>%
  mutate(Rate = Patients / Total_Patients)

# Function to extract numerical part of feature_id for sorting
extract_numeric_part <- function(feature_id) {
  as.numeric(gsub("PheCode:", "", feature_id))
}

# Function to get description for a given PheCode
get_phecode_description <- function(phecode_pattern, dictionary) {
  phecode_pattern <- str_replace(phecode_pattern, "^PheCode:", "")
  description <- dictionary %>%
    filter(feature_id == paste0("PheCode:", phecode_pattern)) %>%
    pull(description)
  
  if (length(description) == 0) {
    return("description not found")  
  } else {
    return(description)
  }
}

# Function to get descriptions for legend labels, ordered numerically
add_description_to_legend <- function(phecode_pattern, dictionary) {
  phecode_pattern <- str_replace(phecode_pattern, "^PheCode:", "")
  dictionary %>%
    filter(str_detect(feature_id, paste0("^PheCode:", phecode_pattern)) & 
           !str_detect(feature_id, "\\.\\d{2,}")) %>%
    mutate(num_part = extract_numeric_part(feature_id)) %>%  # Extract numeric part
    arrange(num_part) %>%  # Order numerically
    mutate(feature_id = paste0(feature_id, " | ", description)) %>%
    dplyr::select(feature_id)
}

# 3.a. Function to plot rate for a PheCode parent-child pair
plot_rate <- function(phecode_pattern, dictionary) {
  phecode_description <- get_phecode_description(phecode_pattern, dictionary)
  legend_labels <- add_description_to_legend(phecode_pattern, dictionary)

  ggplot(phecode_summary %>%
           filter(str_detect(feature_id, paste0("^PheCode:", phecode_pattern)) & 
                  !str_detect(feature_id, "\\.\\d{2,}")),
         aes(x = year, y = Rate, color = feature_id, group = feature_id)) +
    geom_line() +
    geom_point() +
    labs(
      title = "Rates for Parent-Child PheCodes",
      subtitle = paste0("<b><i> PheCode: ", phecode_pattern, "</i></b> | <i>", phecode_description, "</i>"), 
      x = "Year",
      y = "Rate",
      color = "Code"
    ) +
    scale_color_discrete(labels = legend_labels$feature_id) +  
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 10, lineheight = 1.2),  
      legend.text = element_text(size = 6), 
      legend.title = element_text(size = 9)  
    )
}

# 3.b. Function to plot patient counts for PheCode parent-child pair
plot_count <- function(phecode_pattern, dictionary) {
  phecode_description <- get_phecode_description(phecode_pattern, dictionary)
  legend_labels <- add_description_to_legend(phecode_pattern, dictionary)

  ggplot(phecode_summary %>%
           filter(str_detect(feature_id, paste0("^PheCode:", phecode_pattern)) & 
                  !str_detect(feature_id, "\\.\\d{2,}")),
         aes(x = year, y = Patients, color = feature_id, group = feature_id)) +
    geom_line() +
    geom_point() +
    geom_line(
      data = total_patients_per_year,
      aes(x = year, y = Total_Patients, group = 1),
      linetype = "dotted",
      color = "black"
    ) +
    labs(
      title = "Patient Counts for Parent-Child PheCodes",
      subtitle = paste0("<b><i> PheCode: ", phecode_pattern, "</i></b> | <i>", phecode_description, "</i>"), 
      x = "Year",
      y = "Patients",
      color = "Code"
    ) +
    scale_color_discrete(labels = legend_labels$feature_id) +  
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 10, lineheight = 1.2),  
      legend.text = element_text(size = 6),  
      legend.title = element_text(size = 9)  
    )
}

rate_411 <- plot_rate("411", filtered_dictionary) 
count_411 <- plot_count("411", filtered_dictionary) 

rate_250 <- plot_rate("250", filtered_dictionary) 
count_250 <- plot_count("250", filtered_dictionary) 

rate_296 <- plot_rate("296", filtered_dictionary) 
count_296 <- plot_count("296", filtered_dictionary) 

print(rate_411)
print(count_411)
print(rate_250)
print(count_250)
print(rate_296)
print(count_296)

```

*Notes: (1) Rates are calculated as the number of patients with the target PheCode per calendar year divided by the total number of patients with any code in the same year; (2) Black dotted lines represent total patient counts per calendar year (denominator of rate).*

## Module 4: Target PheCode-CUI Trends

This module analyzes the relationship between a target PheCode and CUI by calculating annual rates, patient counts, and intra-patient correlation.

### Outputs

1. Plot of rates over time for target PheCode-CUI pair
2. Plot of patient counts over time for target PheCode-CUI pair
3. Plot of intra-patient correlation over time for target PheCode-CUI pair

```{r module4}

# ----------------------- DO NOT EDIT ----------------------- 

options(repr.plot.width = 8, repr.plot.height = 6)

# Generate data summaries
nlp_target <- nlp[nlp$feature_id == target_cui,]
codified_nlp_target <- rbind(codified,nlp_target)

phecode_cui_wide <- codified_nlp_target %>%
  group_by(year, patient_num, feature_id) %>%
  summarise(Code_Count = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = feature_id,
    values_from = Code_Count,
    values_fill = 0
  )

target_phecode_summary <- codified %>%
  filter(str_detect(feature_id, target_code)) %>%
  group_by(year) %>%
  summarise(
    PheCode_Patients = n_distinct(patient_num),
    .groups = "drop"
  )

target_cui_summary <- nlp %>%
  filter(str_detect(feature_id, target_cui)) %>%
  group_by(year) %>%
  summarise(
    CUI_Patients = n_distinct(patient_num),
    .groups = "drop"
  )

target_total_patients <- codified_nlp_target %>%
  group_by(year) %>%
  summarise(Total_Patients = n_distinct(patient_num), .groups = "drop")

target_combined <- target_phecode_summary %>%
  left_join(target_cui_summary, by = "year") %>%
  left_join(target_total_patients, by = "year") %>%
  mutate(
    PheCode_Rate = PheCode_Patients / Total_Patients,
    CUI_Rate = CUI_Patients / Total_Patients
  )

# Extract descriptions from dictionary
phecode_description <- get_phecode_description(target_code, filtered_dictionary)
cui_description <- get_phecode_description(target_cui, filtered_dictionary)

# 4.a. Plot rate for target PheCode-CUI pair
plot_rates_target <- ggplot(target_combined, aes(x = as.numeric(year))) +
  geom_line(aes(y = PheCode_Rate, color = "Target PheCode", group = 1)) +
  geom_line(aes(y = CUI_Rate, color = "Target CUI", group = 2)) +
  geom_point(aes(y = PheCode_Rate, color = "Target PheCode")) +
  geom_point(aes(y = CUI_Rate, color = "Target CUI")) +
  labs(
    title = paste("Rates for Target Codes"),
    subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                       "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
    x = "Year",
    y = "Rate",
    color = "Code"
  ) +
  theme_minimal() +
  theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 10, lineheight = 1.2),  
      legend.text = element_text(size = 6),  
      legend.title = element_text(size = 9)  
    ) +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 

# 4.b. Plot patient counts for target PheCode-CUI pair
plot_counts_target <- ggplot(target_combined, aes(x = as.numeric(year))) +
  geom_line(aes(y = PheCode_Patients, color = "Target PheCode", group = 1)) +
  geom_line(aes(y = CUI_Patients, color = "Target CUI", group = 2)) +
  geom_point(aes(y = PheCode_Patients, color = "Target PheCode")) +
  geom_point(aes(y = CUI_Patients, color = "Target CUI")) +
  geom_line(
    aes(y = Total_Patients, group = 3),
    linetype = "dotted",
    color = "black"
  ) +  
  labs(
    title = paste("Patient Counts for Target Codes"),
    subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                       "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
    x = "Year",
    y = "Patients",
    color = "Code"
  ) +
  theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 10, lineheight = 1.2),  
      legend.text = element_text(size = 6),  
      legend.title = element_text(size = 9)  
    ) +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 

# 4.c. Function to plot correlation for target PheCode-CUI pair
corr_target <- function(target_code, target_cui) {
  
  # Check if relevant columns exist
  if (!target_code %in% colnames(phecode_cui_wide) || !target_cui %in% colnames(phecode_cui_wide)) {
    stop(paste("Not enough data for Target Code:", target_code, "or Target CUI:", target_cui))
  }
  
  # Calculate correlations by year
  correlation_by_year <- phecode_cui_wide %>%
    group_by(year) %>%
    summarise(
      Correlation = {
        # Get data for the current year
        data_for_year <- pick(everything())
        
        # Check for zero standard deviation
        if (sd(data_for_year[[target_code]], na.rm = TRUE) == 0 ||
            sd(data_for_year[[target_cui]], na.rm = TRUE) == 0) {
          NA  
        } else {
          cor(data_for_year[[target_code]], data_for_year[[target_cui]], use = "pairwise.complete.obs", method = "spearman")
        }
      },
      .groups = "drop"
    ) %>%
    filter(!is.na(Correlation))  # Remove rows with NA correlations
  
  # Plot correlations
  ggplot(correlation_by_year, aes(x = as.numeric(year), y = Correlation)) +
    geom_line(color = "cornflowerblue") +
    geom_point(color = "cornflowerblue") +
    labs(
      title = "Intra-Patient Correlation for Target Codes",
      subtitle = paste0("<i><b>Target PheCode:</b> <i>", target_code, "</i> | <i>", phecode_description, 
                        "<br><b>Target CUI:</b> ", target_cui, "</i> | <i>", cui_description, "</i>"),
      x = "Year",
      y = "Correlation"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 10, lineheight = 1.2),  
      legend.text = element_text(size = 6),  
      legend.title = element_text(size = 9)  
    ) +
  scale_x_continuous(breaks = unique(as.numeric(target_combined$year))) 
}

plot_corr_target <- corr_target(target_code, target_cui)

plot_rates_target
plot_counts_target
plot_corr_target

```

*Notes: (1) Rates are calculated as the number of patients with the target code per calendar year divided by the total number of patients with any code in the same year; (2) Intra-patient correlations are calculated as the Spearman correlation between the code counts for patients in the same year; (3) Black dotted lines represent total patient counts per calendar year (denominator of rate).*

## Module 5: ONCE Feature Trends (Diagnoses, Medications, Labs, Procedures)

This module identifies the top five related codes in four code categories (diagnosis, medication, lab, procedure) based on ONCE feature similarity and tracks their trends over time. It calculates annual rates, patient counts, and their relationship to the target code. 

### Outputs

1. Plot of rates over time for related features
2. Plot of patient counts over time for related features

```{r module5}

# ----------------------- DO NOT EDIT ----------------------- 

options(repr.plot.width = 8, repr.plot.height = 6)  

# Function to extract top 5 variables for each category, ordered by target similarity
get_top_related_codes <- function(keyword, target_code) {
  ONCE_dictionary %>%
    filter(str_detect(feature_id, keyword) & feature_id != target_code) %>%
    filter(feature_id %in% codified$feature_id) %>%  # Ensure the codes exist in codified$feature_id
    arrange(desc(target_similarity)) %>%
    head(5)
}
  
# Function to generate plots for related codes
plot_related_code_trends <- function(Type, target_code) {
  
  type_dict <- list(
    "Diagnosis" = "PheCode",
    "Medication" = "RXNORM",
    "Lab" = "LOINC",
    "Procedure" = "CCS"
  )
  
  if (!(Type %in% names(type_dict))) stop("Invalid Type. Choose from: Diagnoses, Medications, Labs, Procedures")
  
  related_codes <- get_top_related_codes(type_dict[[Type]], target_code)
  
  if (nrow(related_codes) == 0) {
    message(paste("No data on related **", Type, " Codes** in ONCE dictionary.", sep = ""))
    return(NULL)
  }

  selected_vars <- c(target_code, related_codes$feature_id)
  
  related_data <- codified %>%
    filter(feature_id %in% selected_vars) %>%
    group_by(year, feature_id) %>%
    summarise(Patients = n_distinct(patient_num), .groups = "drop")
  
  total_patients <- codified %>%
    group_by(year) %>%
    summarise(Total_Patients = n_distinct(patient_num), .groups = "drop")
  
  combined_data <- related_data %>%
    left_join(total_patients, by = "year") %>%
    mutate(Rate = Patients / Total_Patients)

  # Extract descriptions and target similarity / ensure legend is ordered correctly
  descriptions <- ONCE_dictionary %>%
    filter(feature_id %in% selected_vars) %>%
    dplyr::select(feature_id, description, target_similarity) %>%
    arrange(desc(target_similarity))

  # Merge descriptions and format legend text
  combined_data <- combined_data %>%
    left_join(descriptions, by = "feature_id") %>%
    mutate(description_label = paste0(feature_id, " | ", description, " [", round(target_similarity, 3), "]"))

  target_description <- descriptions$description[descriptions$feature_id == target_code]
  target_description <- ifelse(length(target_description) > 0, target_description, "No Description Available")
  subtitle_text <- paste0("<i><b>Target Code:</b> ", target_code, "</i> | <i>", target_description, "</i>")

  if (all(combined_data$feature_id == target_code)) {
    message(paste("No data on related *", Type, " Codes* in sample.", sep = ""))
    return(NULL)
  }

  plot_theme <- theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, size = 8, hjust = 1),
      plot.subtitle = element_textbox_simple(size = 8, lineheight = 1.2),  
      legend.text = element_text(size = 6),  
      legend.title = element_text(size = 8)
    )

  # Plot rates
  plot_rates <- ggplot(combined_data, aes(x = as.numeric(year), y = Rate, color = reorder(description_label, -target_similarity), group = feature_id)) +
    geom_line(data = combined_data %>% filter(feature_id == target_code), aes(x = as.numeric(year), y = Rate), linetype = "dotted", color = "black") +
    geom_line(data = combined_data %>% filter(feature_id != target_code)) +
    geom_point(data = combined_data %>% filter(feature_id != target_code)) +
    labs(title = paste("Rates for Target Code and Related", Type, "Codes"),
         subtitle = subtitle_text,  
         x = "Year", y = "Rate", color = paste(Type, "Code")) +
    scale_x_continuous(breaks = unique(as.numeric(combined_data$year))) +
    plot_theme

  # Plot patient counts
  plot_counts <- ggplot(combined_data, aes(x = as.numeric(year), y = Patients, color = reorder(description_label, -target_similarity), group = feature_id)) +
    geom_line(data = combined_data %>% filter(feature_id == target_code), aes(x = as.numeric(year), y = Patients), linetype = "dotted", color = "black") +
    geom_line(data = combined_data %>% filter(feature_id != target_code)) +
    geom_point(data = combined_data %>% filter(feature_id != target_code)) +
    labs(title = paste("Patient Counts for Target Code and Related", Type, "Codes"),
         subtitle = subtitle_text,  
         x = "Year", y = "Patients", color = paste(Type, "Code")) +
    scale_x_continuous(breaks = unique(as.numeric(combined_data$year))) +
    plot_theme

  return(list(plot_rates, plot_counts))
}

# Function to generate and print plots
generate_plots <- function(types, target_code) {
  for (Type in types) {
    plots <- plot_related_code_trends(Type, target_code)
    if (!is.null(plots)) {
      print(plots[[1]])  
      print(plots[[2]])  
    }
  }
}

generate_plots(c("Diagnosis", "Medication", "Lab", "Procedure"), target_code)

```

*Notes: (1) Black dotted lines represent rates and counts for the target code per calendar year.*


